---
title: |
  <span style="color: blue;">Spatial Statistics in Epidemiology and Public Health - SISMID 2026</span><br>
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    html_document:
      toc: true       
      toc_depth: 2
      number_sections: true
      code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
h1 { font-size: 24px; } /* # Heading 1 */
h2 { font-size: 20px; } /* ## Heading 2 */
h3 { font-size: 16px; } /* ### Heading 3 */
</style>
```

<br>

# Mini Project Overview

In this lab session, we will work through the following key steps in
conducting a spatial analysis.

1.  Identify and acquire geospatial datasets.
2.  Prepare spatial data objects in R.
3.  Visualize the data with maps.
4.  Perform exploratory data analysis.
5.  Specify and fit regression models.
6.  Summarize results.

Utilizing the skillsets we just learned, this lab session will focus on
analyzing <u>**county-level data**</u>. You are encouraged to come up
with a unique mini research question by identifying an outcome and a
risk factor of interest using the datasets describe below. You may also
choose to just follow along with the examples!

<br> <br>

# Finding Geospatial Datasets

**Suggested Time: 20 minutes**

Publicly available data products often have an online visualization tool
to help users explore data elements and assist with data download. Below
are three example products by the US CDC. Spend a few minutes exploring
the variable available at the county-level.

1.  **AtlasPlus** from the National Center for HIV, Viral Hepatitis,
    STD, and Tuberculosis Prevention.
    - Link: <https://gis.cdc.gov/grasp/nchhstpatlas/maps.html>
    - About: <https://www.cdc.gov/nchhstp/about/atlasplus.html>
2.  **Data Explorer** from the National Environmental Public Heatlh
    Tracking Network
    - Link: <https://ephtracking.cdc.gov/DataExplorer/>
    - About: <https://ephtracking.cdc.gov/indicatorPages>

<br>\

::: {style="background-color: #f4f6f8; border: 1px solid #ccd1d9; border-left: 6px solid #0076b6; padding: 8px; border-radius: 4px;"}
<h4>**Specific Tasks**</h4>

- Identify your research question by picking an outcome of interest and
  some covariates with data available at the county-level.
- Select a study region with at least 50 counties (e.g., a single state
  of a few contiguous states).
- Figure out how to download the data (typically in a .csv file format).
- Share in the Slack Channel about what research topic you have chosen!
:::

<br>

<p>

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#collapseBox" aria-expanded="false" aria-controls="collapseBox">

Example

</button>

</p>

:::: {#collapseBox .collapse}
::: {.card .card-body style="background-color: #f8f9fa; border: 1px solid #ddd; padding: 15px;"}
This analysis aims to estimate associations between emergency department
visits for asthma and particulate matter air pollution from diesel. We
will utilize the 2019 annual county-level data from the National
Environmental Public Health Tracking Network.The study region will
include 4 states from the Mid-Atlantic and Northeast region (PA, MD, NY
and NJ). We also obtained variables on the population size.

The downloaded datasets *asthmaED_2019.csv*, *diesel_2019.csv* and
*population_2019.csv* are on the course website.
:::
::::

<br> <br>

# Data Wrangling

**Suggested Time: 20 minutes**

The first step is to construct an analytic data from the files we just
downloaded. We will use the county 5-digit ID (FIPS Code) for linkage.
Here are the issues we encountered and some potential other ones.

- Missing values are in given text "suppressed," which we re-code to NA
  and convert the variable to numeric.\
- Population size is stratified by age group, and we sum up all groups
  within each county.
- Some county FIPS ID has less than 5 digits because of missing leading
  zeros, which needs to be padded.

<br>\

::: {style="background-color: #f4f6f8; border: 1px solid #ccd1d9; border-left: 6px solid #0076b6; padding: 8px; border-radius: 4px;"}
<h4>**Specific Tasks**</h4>

- Identify the variables needed from the downloaded datasets.
- Check issues related to missing data.
- Examine summary statistics of the variables and create some
  exploratory plots (boxplot, histogram, scatter plot).
:::

<br>

<p>

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#collapseBox2" aria-expanded="false" aria-controls="collapseBox">

Example

</button>

</p>

:::: {#collapseBox2 .collapse}
::: {.card .card-body style="background-color: #f8f9fa; border: 1px solid #ddd; padding: 15px;"}
```{r, message = FALSE, warning = FALSE}
install.packages("pacman")

pacman::p_load(patchwork,
               dplyr,
               readr,
               table1,
               here)
library (dplyr)
library (readr)
library (table1)

here::i_am("Lab 1.Rmd")

### Read in CSV files ###
asthma = read.csv ("Lab Example Data/asthmaED_2019.csv")
diesel = read.csv ("Lab Example Data/diesel_2019.csv")
pop = read.csv ("Lab Example Data/population_2019.csv")

### Rename variable
asthma =  rename (asthma, asthma = Value)
diesel = rename (diesel, diesel = Value)

#Replace suppressed values with NA
asthma = asthma %>%
  mutate (asthma = na_if (asthma, "Suppressed"))
asthma$asthma = parse_number(asthma$asthma)

#Aggregate age-group population to total population
pop$pop = parse_number(pop$Value)
pop_tot = pop %>% 
  group_by (CountyFIPS) %>% 
  summarize (pop_total = sum (pop))

# Subset to study states ###
dat = subset (asthma, State %in% c("Maryland", "Pennsylvania", "New York", "New Jersey") )
dat =  dat %>% select (State, CountyFIPS, County,  asthma)

# Merge data by CountyFIPS
dat = dat %>% 
  left_join(diesel %>% select (CountyFIPS, diesel), by = join_by (CountyFIPS)) %>%
  left_join(pop_tot, by = join_by (CountyFIPS))

dat$rate = round(dat$asthma/dat$pop_total*10000 )

table1(~State + asthma+ diesel + pop_total + rate, data = dat)
```

```{r, fig.align="center", out.width="60%"}
plot (dat$rate~dat$diesel, xlab = "Annual Diesel PM Pollution Concentration", ylab = "Astham ED Rate (per 10,000)", main= "Annual Asthma Emergency Department Visit Rates in 2019")
```
:::
::::

<br> <br>

# Creating Geospatial Datasets and Mapping

**Suggested Time: 20 minutes**

For area-level data, we need to associate geospatial polygon information
to each county. This is often included in a *shapefiles*. The US Census
Bureau [publishes
shapefiles](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.2024.html#list-tab-790442341)
for various geographica units (e.g., county, tracts, ZCTA). Here we will
use *tigris* package that downloads from the Census. Below is the code
snippet for our example with 4 states.

```{r, message = FALSE, warning = FALSE, class.source = "fold-show"}

library (tigris)
options(tigris_use_cache = TRUE)

#option cb = TRUE to download the coarser version for speed
#we choose the year 2020
dat_sf <- counties(state = c("MD", "PA", "NY", "NJ"), cb = TRUE, year = 2020)
```

::: {style="background-color: #f4f6f8; border: 1px solid #ccd1d9; border-left: 6px solid #0076b6; padding: 8px; border-radius: 4px;"}
<h4>**Specific Tasks**</h4>

- Download the county shapefile for your study region.
- Create two county maps showing your outcome and exposure of interests.
  Share on Slack channel!
:::

<br>

<p>

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#collapseBox4" aria-expanded="false" aria-controls="collapseBox">

Example

</button>

</p>

:::: {#collapseBox4 .collapse}
::: {.card .card-body style="background-color: #f8f9fa; border: 1px solid #ddd; padding: 15px;"}
Note that the resulting dataset is an *sf* object. Geometry is recorded
in latitude and longtidue (NAD83). The dataset already contains several
variables. **GEOID** is the 5-digit FIPS code for linking to other
datasets. Below are the first 3 rows of the download county shapefil
dataset and its meta data. A quick map shows that we have acquired the
correct shapefile!

```{r}
head (dat_sf, 3)
```

```{r, fig.align="center", out.width="60%"}
library (ggplot2)
ggplot(data = dat_sf) +
  geom_sf(fill = "whitesmoke", color = "gray60", size = 0.1) +  theme_void() +
  labs(title = "Study Region")
```

Next, we merge our dataset with the county sf object, and create
choropleth maps for the exposure and outcome. *dplyr* joint functions
can work with *sf*. Note that two PA counties with missing ED data in
grey.

```{r, fig.align="center", out.width="90%"}
library (patchwork)
dat$CountyFIPS = as.character (dat$CountyFIPS)
dat_sf = dat_sf %>% left_join (dat,  by = join_by (GEOID == CountyFIPS) )

g1 = ggplot(data = dat_sf) +
  geom_sf(aes(fill = rate)) + 
  labs(title="Asthma ED Rates per 10,000")+scale_fill_gradient2(low = "white",high = "blue")

g2= ggplot(data = dat_sf) +
  geom_sf(aes(fill = diesel)) + 
  labs(title="Diesel Particulate Matter Pollution")+scale_fill_gradient2(low = "white",high = "red")

g1+g2

```
:::
::::

<br> <br>

# Spatial Exploratory Analysis

**Suggested Time: 20 minutes**

Based the outcome, study region and maps from Section 4, what should we
use as the neighbor/network structure to induce spatial correlation? The
most commonly used one for county-level is to use whether two counties
share a border.

<br>

::: {style="background-color: #f4f6f8; border: 1px solid #ccd1d9; border-left: 6px solid #0076b6; padding: 8px; border-radius: 4px;"}
<h4>**Specific Tasks**</h4>

- Construct an adjacency matrix for the analysis.
- Is there evidence of spatial auto-correlation in the outcome and
  covariates?
:::

<br>

<p>

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#collapseBox5" aria-expanded="false" aria-controls="collapseBox">

Example

</button>

</p>

:::: {#collapseBox5 .collapse}
::: {.card .card-body style="background-color: #f8f9fa; border: 1px solid #ddd; padding: 15px;"}
In our example, we define spatial neighbors using *either* the 1st-order
adjacency (i.e., sharing a boundary) or two counties have centroids that
are within 25 km. This is because some small counties counties do not
share a border across rivers. We also drop the two counties with missing
data.

```{r, message = FALSE, warning = FALSE, class.source = "fold-show"}
library (spdep)

#Drop missing data
dat_sf = subset (dat_sf, !is.na (dat_sf$rate))

#Project the data to, for example, the Lambert Conformal Conic North America.
dat_proj = st_transform(dat_sf, crs = "ESRI:102004")
center_proj = st_centroid(dat_proj)

#Create 1-st order adjacency and distance buffer (100km)
nb.1st = poly2nb (dat_proj)
nb.buffer = dnearneigh(center_proj, d1=0, d2 = 25*1000)
nb = union.nb (nb.1st, nb.buffer) #take the union of neighbors

#Create weight matrix
W= nb2listw(nb, style="W")
W

#Monte Carlo-based Moran's I
moran.mc(dat_sf$rate, W, nsim = 10000)
moran.mc(dat_sf$diesel, W, nsim = 10000)

```

<br>

**Conclusion**: We observe positive Moran's I for both asthma ED rates
(0.29) and diesel air pollution (0.86) with p-value \< 0.01 based a
permutation test of 10,000 Monte Carlo simulations against the null
hypothesis of no spatial autocorrelation. We can also see from the maps
from Section 4 that exposure variable is more spatially homogeneous.
:::
::::

<br> <br>

# Spatial Regression Modeling

**Suggested Time: 20 minutes**

<br> We will now use conditional autoregressive models to estimate the
association between the outcome and covariates.

<br>

::: {style="background-color: #f4f6f8; border: 1px solid #ccd1d9; border-left: 6px solid #0076b6; padding: 8px; border-radius: 4px;"}
<h4>**Specific Tasks**</h4>

- Describe or write down the statistical model.
- Fit the model using INLA and summarize results.
- Compare your model to a conventional model without accounting for
  spatial dependence.
:::

<br>

<p>

<button class="btn btn-primary" type="button" data-toggle="collapse" data-target="#collapseBox6" aria-expanded="false" aria-controls="collapseBox">

Example

</button>

</p>

:::: {#collapseBox6 .collapse}
::: {.card .card-body style="background-color: #f8f9fa; border: 1px solid #ddd; padding: 15px;"}
For county $s$, let $Y_s$ be the number of asthma ED visit, $X_s$ be the
diesal air pollution exposure, and $P_s$ be the population size. We
consider several Poisson log-linear models of the following form.

$$\begin{aligned}
Y_s &\sim \operatorname{Poisson}(\mu_s) \\
\log(\mu_s) &= \log(P_s) + \beta_0 + \beta_1 X_{s} + \theta_s
\end{aligned}$$

1.  The standard Poisson model assumes no spatial variation
    ($\theta_s = 0$ for all $s$).

2.  The independent random intercept model assumes that
    $\theta_s \sim N(0, \tau^2)$.

3.  The spatial random intercept model (under the improper CAR model
    formulation) assumes that $\theta_s = \eta_s + \gamma_s$, where
    $\eta_s \sim N(0, \tau^2)$ and $\gamma_s$ is a conditional
    autoregressive model.

```{r, message = FALSE, warning = FALSE}
install.packages("INLA",
  repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/testing"),
  type = "binary",
  dep = TRUE)
library (INLA)

#Some data wrangling first
dat_sf$areal_ID = 1:nrow (dat_sf) #Create a county ID from 1 to 226
dat_sf$log_pop = log (dat_sf$pop_total)
dat_sf$diesel_std = dat_sf$diesel / 0.1 #standardized by the interquartile range (0.1)

#INLA needs a specific summary of adjacency structure
nb2INLA ("adj.txt", nb) #write out a file
G <- inla.read.graph(filename = "adj.txt") #read in to get INLA's graph format

#Fit glm
fit.glm = glm (asthma~diesel_std+offset (log_pop), dat = dat_sf, family = "poisson")

#Fit exchangeable (random effect model)
#Note: "E" here is the offset and it does not need to be log-transformed
fit.exch = inla (asthma~1+f(areal_ID)+diesel_std, E= pop_total, family = "poisson", data = dat_sf,
                 control.compute = list(dic = TRUE, waic = TRUE, return.marginals.predictor=TRUE))

#Fit improper CAR model (Besag)
fit.iCAR = inla (asthma~1+f(areal_ID,  model = "besag", graph = G)+diesel_std, E= pop_total, family = "poisson", data = dat_sf,control.compute = list(dic = TRUE, waic = TRUE))

```

The table below summarizes the estimated relative rate $e^{\beta_1}$ per
interquartile range increase in diesel air pollution.

```{r}

### Extract coefficients from each model
library (knitr)
glm.coef = summary(fit.glm)$coefficients
fit.iCAR.coef = fit.iCAR$summary.fixed
fit.exch.coef = fit.exch$summary.fixed

coefs = round(c(glm.coef[2,1], fit.exch.coef$mean[2], fit.iCAR.coef$mean[2]),2)
se =  round(c(glm.coef[2,2], fit.exch.coef$sd[2], fit.iCAR.coef$sd[2]),2)
low95 = round(exp(c( coefs[1]-1.96*se[1], fit.exch.coef$`0.025quant`[2], fit.iCAR.coef$`0.025quant`[2])), 2)
up95 = round(exp(c( coefs[1]+1.96*se[1], fit.exch.coef$`0.975quant`[2], fit.iCAR.coef$`0.975quant`[2])),2)
results = data.frame (
  model = c("GLM", "Ind Random Effect", "Spatial"),
  beta = coefs, se = se, rr = round(exp (coefs),2), ci = paste0("(", low95, ", ", up95, ")"))

kable (results,
       col.names = c("Model", "&beta;", "Std Err", "RR", "95% Interval"))

```

- All 3 models show the counties with higher level of diesel pollution
  are associated with with higher rate of asthma ED visits.
- The model without accounting for spatial correlation have the smallest
  standard error, which is likely an under-estimation.
- The spatial model estimates that a 21% increase in asthma ED visit is
  associated per IQR range increase in exposure.
:::
::::

<br> <br> <br>
